---
title: "amplican overview"
author: "Kornel Labun & Eivind Valen"
date: "`r doc_date()`"
package: "`r pkg_ver('amplican')`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{amplican overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Welcome to the `amplican` package. This vignette will walk you through our main package usage with example MiSeq 
dataset.  You will learn how to interpret results, create summary reports and plot deletions, 
insertions and mutations with our functions. 
This package, `amplican`, is for fast and precise analysis of CRISPR experiments.

# Introduction

`amplican` creates reports of deletions, insertions, frameshifts, cut rates and other metrics in *knitable* to HTML style. `amplican` uses vary fast C implementation of Gotoh algorithm to align your fastq samples and
automates analysis across different experiments. `amplican` maintains elasticity through configuration file, which with your fastq samples are only requirements.

For those inpatient people who want to see an example of our whole pipeline analysis on our example data start [here](#default-options).

# Configuration file

To successfully run our analysis it is mandatory to have a configuration file. Take a look at our example:

```{r, echo=FALSE}
config <- read.csv(system.file("extdata", "config.csv", package = "amplican"))
knitr::kable(head(config))
```

Configuration file should be a "," delimited csv file with information about your experiments. 
You can find example config file path by running: 
```{r configPath, echo=TRUE, eval=FALSE}
system.file("extdata", "config.csv", package = "amplican")
```

Columns of the config file:  
**ID** - should be a unique identifier of your experiments  
**Barcode** - use this to group experiments with the same barcode  
**Forward_Reads**, **Reverse_Reads** - put names of your files in these fields, you can put here files compressed to 
.gz, we will unpack them for you  
**Group** - use this field if you want to group ID by other criteria, here for instance we will group by person that
performed experiment, on later stage it is possible to generate report with breakdown using this field  
**guideRNA** - put your guideRNA here, if he has to be reverse complemented to be found in the amplicon put "1" into **Direction** field, we will use this field to make sure your guideRNA is inside the amplicon  
**Forward_Primer**, **Reverse_Primer** - make sure these primers are correct for each of your IDs  
**Direction** - here "0" means guide does not requires to be reverse complemented to be matched to the amplicon, 
"1" means it should be reverse complemented, this field is also used when plotting deletions, mismatches and 
insertions  
**Amplicon** - insert amplicon sequence as lower case letters, by putting UPPER CASE letters you can specify
expected cut site, you should specify at least one cut site, here for example we specify in uppercase letter PAM 
sequence of the corresponding guideRNA  

# Default options

To run `amplican` with default options, along with generation of all posible reports you can use
`amplicanPipeline` function. We have already attached results of the default amplican analysis of the
example dataset, but take a look how you can do that yourself. Be prepared to grab a coffe when running
`amplicanPipeline` with `knit_files = TRUE` as this will take some time. You will understand it is worth waiting when reports will be ready. We don't ship with these reports as they take up too much space to be included in R package.

```{r amplicanPipeline, echo=TRUE, eval=FALSE}
# example MiSeq dataset files
config <- system.file("extdata", "config.csv", package = "amplican") #path to example config file
fastq_folder <- system.file("extdata", "", package = "amplican") #path to example fastq files
results_folder <- paste0(getwd(), "/amplican_example") #output folder

## make sure we have results folder
dir.create(results_folder)
##  actual amplican run
amplicanPipeline(config, fastq_folder, results_folder)

# results of the analysis can be found at
message(results_folder)
```

Take a look into "results_folder" folder. Here you can find `.Rmd` files that are our reports for example dataset. We already crafted `.html` versions and you can find them in the "html" folder. Open one of the reports with your favourite browser now. To zoom on an image just open it in new window (right mouse click options). 

`amplicanPipeline` just crafted very detailed reports for you, but this is not all, if you need something different e.g. different plot colours, just change the `.Rmd` file and `knit` it again. This way you have all the power over plotting.

# Files created during analysis

## alignmentLog.txt

Contains information about unpleasant errors and events found during analysis, we also print message on the screen when something requires your attention. This file should be empty with our example analysis on provided dataset. In case something goes wrong you should examine file and follow instructions inside. This file is also included in "summary" version of the report.

## barcode_reads_filters.csv

First step of our analysis is to filter out reads that are not complying with our default restrictions:  

* bad base quality - default minimum base quality is 0  
* bad average quality - default minimum average quality is 0  
* bad alphabet - by default we accept only reads with A,C,T,G bases  

```{r, echo=FALSE}
barcodeFilters <- read.csv(system.file("extdata/results", "barcode_reads_filters.csv", package = "amplican"))
knitr::kable(head(barcodeFilters))
```

This table is also summarized in one of the reports. As you can see for our example dataset we have two barcodes, to which correspond 17 and 29 reads. Two reads are rejected for barcode_1 due to bad alphabet. Each of the barcodes has unique reads, which means forward and reverse reads are compacted when they are identical. There is 11 unique reads for barcode_1 and 20 for barcode_2. We can successfully assign all the reads for each barcode as this is artificial dataset. Normally you will probably see only half of your reads being assigned to the barcodes. Reads are assigned when for forward read we can find forward primer and for reverse read we can find reverse primer. Primers have to be perfectly matched. Nevertheless, there is option `fastqreads = 0.5` which changes method of assigning reads to each IDs. With this option specified only one of the reads (forward or reverse) have to have primer perfectly matched.

## config_summary.csv

`config_summary.csv` contains extended version of the config file. It should provide you additional look at raw numbers which we use for various plots in reports. Take a look at example extension:

```{r, echo=FALSE}
config_summary <- read.csv(system.file("extdata/results", "config_summary.csv", package = "amplican"))
config_summary <- config_summary[,c("ID", "Barcode", "Cut", "Frameshift", 
                                    "PRIMER_DIMER", "Reads", "Found_Guide", "Found_PAM")]
knitr::kable(head(config_summary))
```

During `amplicanAlign` or `amplicanPipeline` these columns are added to the config file:  
**Cut** - number of insertions overlapping with user specified UPPER CASE group in the amplicon (extended by the buffer)  
**Frameshift** - number of reads has frameshift (insertions and deletions)  
**PRIMER_DIMER** - number of reads were classified as PRIMER DIMERs  
**Reads** - number of reads assigned to this ID  
**Found_Guide** - whether we found guide on the amplicon  
**Found_PAM** - whether amplicon has UPPER GROUP specified (advised)  

## RunParameters.txt

File RunParameters.txt lists all options used for the analysis, this file you might find useful when 
reviewing analysis from the past where you forgot what kind of options you used. Other than that 
this file has no purpose. 

```{r, echo=TRUE}
readLines(system.file("extdata/results", "RunParameters.txt", package = "amplican")) #path to example RunParameters.txt
```

## "alignments" folder

Contains alignments, for each ID there is separate folder. If you specified `write_alignments` option of the 
`amplicanAlign` as "2", which is default, you will find here two files, one with just alignment forward read, 
reverse read and count.
Other file will contain very detailed information:  
- human readable alignment of read and amplicon,  
- list of deletions, insertions and mismatches,  
- same information, but relative towards amplicon coordinates

```{r, echo=TRUE, eval=FALSE}
# path to the example alignments folder
system.file("extdata", "results/alignments", package = "amplican")
```

In `unassigned_sequences` folder you can find detailed information about unassigned reads. In example dataset there is no unassigned reads.

Take a look at the alignment events file which contains all the insertions, deletions, cuts and mutations. This file can be used in various ways. Examples you can find in `.Rmd` files we prepare using `amplicanReport`. These can be easily converted into `GRanges` and used for further analysis!
```{r, echo=FALSE}
alignments <- read.csv(system.file("extdata", "results/alignments.csv", package = "amplican"))
knitr::kable(head(alignments))
```

Human readable alignments format gives information about forward and reverse reads. First line is corresponding to the forward alignment and second line is always for the reverse read alignment. This alignment does not shows amplicon, just similarity between two alignments (forward and reverse). Human readable format looks like this:
```{r, echo=FALSE}
readLines(system.file("extdata", "results/alignments/ID_1_barcode_1/alignments.txt", package = "amplican"), n = 6)
```


Human readable detailed alignments format gives more information. To make more detailed alignment files `amplicanPipeline` or `amplicanAlign` should have option `write_alignments == 2` specified (Default behavior), but it adds a little to the computational time.
```{r, echo=FALSE}
readLines(system.file("extdata", "results/alignments/ID_1_barcode_1/detailed_alignments.txt", package = "amplican"), n = 89)
```


## reports folder

Reports are automated for the convenience of the users. We provide 6 different reports. Reports are .Rmd files which can be easily crafted through `rmarkdown::render(path_to_report)` or by clicking `Knit` in Rstudio to make HTML version of the report. If you have run our example analysis, then you can open one of the reports with Rstudio and  try knitting it now! Otherwise you should change the path at the top of this file to match your path to the package, as you have pre-generated path (running `amplicanPipeline` does that for you). 

```{r, echo=TRUE, eval=FALSE}
# path to example reports folder
system.file("extdata", "results/reports", package = "amplican") #folder with example reports

# you can run code below as example how to knit file:
rmarkdown::render(system.file("extdata", "results/reports/report_id.Rmd", package = "amplican"))
```

# Detailed analysis

## Aligning reads

Each of the reads, that is assigned to unique ID, is aligned with usage of the Gotoh algorithm 
implemented in C for speed. When you want to have more control over alignments and you need more advanced options use `amplicanAlign`. This function has many parameters which can be used to bend our pipeline to your needs. 
Read more about `amplicanAlign` on the help page `?amplican::amplicanAlign`.

## Making reports

Reports are made for user convenience, they are powerful as they:

* contain reproducible code
* make plots ready for publication
* show how to use results of our pipeline
* are easy to share
* allow quick, but detailed assessment of the data

We decided to separate reports into 6 different types. Function `amplicanReport` is made for automated report creation.

Types of report:

* **ID** - Each of the IDs is treated separately, you get to see what kind of cuts, deletions, insertions and mismatches
are being found for each of the IDs.
* **Amplicon** - We aggregate information about each unique amplicon, you will find here also plots for events found 
during alignment.
* **Barcode** - Grouped by Barcode, instead of the information about alignment events you can find here histograms 
explaining cut rates, frameshifts etc.
* **Group** - This field is for your convenience and allows you to group IDs by anything you want. In our example dataset
we use this field to group experiments by the users. That will allow us to asses performance of our technicians.
* **Guide** - Unique guideRNAs are being summarized by boxplots showing cut rates across all experiments. 
* **Summary** - Contains overall statistics gathered during reads assesment. Can contain only barcode level
information, but shows which parts of reads is being filtered.

## Plotting alignments events

We provide specialized plots for each type of the alignment events. 

`amplican_plot_mismatches` - plots mismatches as stacked bar-plot over the amplicon, top plot is for the forward and
bottom is for reverse reads  
`amplican_plot_deletions` - plots deletions as arch-plot from the ggbio package  
`amplican_plot_insertions` - each insertion is represented by triangle over the amplicon  
`amplican_plot_cuts` - gathers cuts from each of the experiments and overlays multiple arch-plots on top of each other,
useful when analyzing what kind of cuts were introduced in the amplicon  

You can take a look at all these plots by *knitting* amplicon report file located here:

```{r, echo=TRUE, eval=FALSE}
# path to example knitable report
system.file("extdata", "results/reports/report_amplicon.Rmd", package = "amplican")
```
