---
title: "amplican overview"
author: "Kornel Labun"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Welcome to the `amplican` package. This viginnette will walk you through our main package usage with example MiSeq dataset. 
You will learn how to interpret results, create summary reports and plot deletions, insertions and mutations with
our functions. This package, `amplican` is aimed into analysis of MiSeq or HiSeq PCR data, whether paired-end or 
single-end. 

## Introduction

`amplican` creates raports of deletions, insertions, frameshifts, cut rates and other metrics in user selected format 
(preffered html). `amplican` uses vary fast C implementation of Gotoh alhoritm to align your fastq samples and 
automates analysis across different experiments. `amplican` maintains elasticity through configuration file, 
which with your fastq samples are only requirements.

## Configuration file

To succesfully run our analysis it is mandatory to have created configuration file. Take a look at our example:

```{r, echo=FALSE, results = 'asis'}
config <- read.csv(system.file("extdata", "config.txt", package = "amplican"), sep = "\t")

knitr::kable(head(config))
```

Configuration file should be a tab delimited text file with information about your experiments. 
You can find example config file at:
```{r, echo=FALSE}
system.file("extdata", "config.txt", package = "amplican")
```

## Default options

The figure sizes have been customised so that you can easily put two images side-by-side.

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## Detailed analysis

### Aligning reads

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

### Printing reports
Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
